numeric vs HPO

### 2022-04-26 code/r/02_D_numeric_vs_hpo_for_CRC.Rmd
continue refactoring

if (!dir.exists(CACHE_DIR)){
  dir.create(CACHE_DIR)
}
FORCE_RERUN <- TRUE
```

@@

ggsave(filename = 'cache/CRC_labs_raw_data_distribution_by_rel_date.png', plot = p, width = 8, height = 6)
```

![liver function labs](./cache/CRC_labs_raw_data_distribution_by_rel_date.png)

Use -180 as the time cutoff and separate the patient journey into two sections: 0 - 180 days before the first diagnosis of abnormal liver function, and 180 days before the first diagnosis of liver abnormal functions. Within each section, we will pick the latest lab measurement for each LOINC lab test (features) and assign either + (within 180 days) or - outcome (outside of 180 days). 
```{r}
# prepare data entries for + and - outcomes
# each lab result is associated with an outcome
# briefly, the following snippet uses the following algorithm:
# calculate the relative date to the first abnormal liver function diagnosis: if before diagnosis, - days otherwise + days
# then cut the relative days into a few windows (-Inf, -365), (-365, -30), (-30, -7), (-7, 0) and (0, Inf)
# for each window, we only use the closest one to diagnosis date as a representative, although we can use other methods such as the mean
# at last, only the (-7, 0) window has + outcome, meaning having abnormal liver function; all other windows have normal liver function

@@

```{r max.output=-1, message=FALSE}
require(furrr)
plan(multisession, workers = 5)

crc_analysis <- df_labs_and_outcomes %>%
  group_by(loinc) %>%
  nest() %>%
  ungroup() %>%
  mutate(result = future_map(data, pipeline_compare_for_outcome_prediction)) %>%
  select(-data) %>%
  unnest(result)

crc_analysis
```

```{r message=FALSE}
df_one_loinc <- df_labs_and_outcomes %>% filter(loinc == '4544-3')

df_hpo_features <- create_hpo_features(df_one_loinc = df_one_loinc)

df_train_test <- train_test_split(df_hpo_features, p = 0.7, list=FALSE)

lr_L <- lr_for_hpo_L(df_train = df_train, df_test = df_test)
roc_comparison <- compare_rocs(lr_L[[5]], lr_L[[6]])
roc_comparison <- bind_cols(roc_comparison)

plot(lr_L[[5]])
plot(lr_L[[6]])

roc.test(lr_L[[5]], lr_L[[6]], method = "bootstrap")
```

@@

```{r max.output=-1, message=FALSE}
require(furrr)
plan(multisession, workers=8)
start_time <- Sys.time()

crc_separate_controls_analysis <- df_labs_and_outcomes_case_control %>%
  group_by(loinc) %>%
  nest() %>%
  ungroup() %>%
  mutate(result = future_map(data, pipeline_compare_for_outcome_prediction, .options = furrr_options(seed = T))) %>%
  select(-data) %>%
  unnest(result)

crc_separate_controls_analysis
```

```{r message=FALSE}
df_one_loinc <- df_labs_and_outcomes_case_control %>% filter(loinc == '4544-3')

df_hpo_features <- create_hpo_features(df_one_loinc = df_one_loinc)

df_train_test <- train_test_split(df_hpo_features, p = 0.7, list=FALSE)

lr_L <- lr_for_hpo_L(df_train = df_train, df_test = df_test)
roc_comparison <- compare_rocs(lr_L[[5]], lr_L[[6]])
roc_comparison <- bind_cols(roc_comparison)

plot(lr_L[[5]])
plot(lr_L[[6]])

roc.test(lr_L[[5]], lr_L[[6]], method = "bootstrap")
```

@@

ggplot(df_demographics_with_outcomes, aes(x = outcome, y = propensity_score)) +
  geom_violin()
```
```{r}
require(MatchIt)

@@

# try this stackoverflow method
# https://stackoverflow.com/questions/29672088/speedup-matchit
# still slow, but works
df_demographics_with_outcomes %>% colnames()
m.out <- matchit(outcome ~ medical_record_number, data = df_demographics_with_outcomes, method = "nearest", distance = df_demographics_with_outcomes$propensity_score)
save(m.out, file='m.out.obj')
matched_unique_ids_etc <- match.data(m.out, data = df_demographics_with_outcomes)

matched_unique_ids_etc %>% group_by(outcome) %>% summarise(n =n())
ggplot(matched_unique_ids_etc, aes(x = outcome, y = propensity_score)) +
  geom_violin()

plot_match_age <- ggplot(matched_unique_ids_etc, aes(x = outcome, y = age, fill = outcome)) +
  geom_violin(draw_quantiles = c(0.5)) +
  theme(legend.position = 'none')

plot_match_gender <- ggplot(matched_unique_ids_etc, aes(x = outcome, fill = gender)) +
  geom_bar(stat = 'count')

plot_match_race <- ggplot(matched_unique_ids_etc, aes(x = outcome, fill = race)) +
  geom_bar(stat = 'count')

plot_match <- gridExtra::arrangeGrob(plot_match_age, plot_match_gender, plot_match_race, layout_matrix = rbind(c(1,2), c(3,3)), ncol = 1)
plot(plot_match)
```
We need to put back the identifiers of controls to the database in order to selectively retrieve their labs. 
```{r}
dbExecute(dsca, "create table if not exists matched_control_cohort (medical_record_number VARCHAR)")
dbExecute(dsca, "truncate table matched_control_cohort")

@@

matched_unique_ids_etc %>% filter(outcome == 'normal') %>% 
  rowwise() %>%
  mutate(batch_number = sample(seq(10), size = 1, replace = TRUE)) %>%
  ungroup() %>%
  group_by(batch_number) %>%
  summarise(sql_query = str_c(str_c("('", medical_record_number, "')"), collapse = ",")) %>% 
  ungroup() %>%
  mutate(sql_query = walk(sql_query, insert_mrns, dbconn = dsca))

# retrieve labs for the matched control cohort
sql <- read_file('../sql/get_labs_for_colorectal_cancer_matched_control.sql')
df_matched_control_labs <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = 'labs_colorectal_cancer_matched_controls.csv', db_access_callback = dbGetQuery, conn = dsca, statement = sql, force_rerun = TRUE)

# drop the matched control cohort
# persist it for a while for now
# n <- dbExecute(dsca, "DROP TABLE IF EXISTS matched_control_cohort")
# assertthat::are_equal(n, 0)
```

@@

```{r max.output=-1, message=FALSE}
require(furrr)
plan(multisession, workers=4)
start_time <- Sys.time()

@@

crc_case_matched_controls_analysis %>% write.csv(file = 'cache/crc_case_matched_controls_analysis.csv', row.names = FALSE)

crc_case_matched_controls_analysis

```

Combine demographics to see whether the AUC can be improved. 
```{r}
df_demographics_case_matched_control <- bind_rows(df_case_demographics %>% select(-min_diag_date), 
                                                  df_control_demographics %>% filter(medical_record_number %in% 
                                                                                       unique(df_matched_control_labs$medical_record_number)) %>%
                                                    filter(!(medical_record_number %in% df_case_demographics$medical_record_number)))

plan(multisession, workers=4)
start_time <- Sys.time()

crc_case_matched_controls_analysis_with_demographics <- df_labs_and_outcomes_case_matched_control %>%
  left_join(df_demographics_case_matched_control, by = 'medical_record_number') %>%
  mutate(age = as.duration(interval(date_of_birth, lab_date)) / dyears()) %>%
  filter(!(race %in% c('African American', 'American Indian or Alaska Native', 'Native Hawaiian or Other Pacific Islander'))) %>%
  # mutate(gender = as.factor(gender),
  #        race = as.factor(race)) %>% 
  # mutate(gender = relevel(gender, 'Male'),
  #        race = relevel(race, 'White')) %>%
  group_by(loinc) %>%
  nest() %>%
  ungroup() %>% 
  mutate(result = future_map(data, pipeline_compare_for_outcome_prediction, include_demographics=TRUE, .options = furrr_options(seed = T))) %>%
  select(-data) %>%
  unnest(result)

crc_case_matched_controls_analysis_with_demographics
```

>>>>

if (!dir.exists(CACHE_DIR)){
  dir.create(CACHE_DIR)
}
FORCE_RERUN <- FALSE
```

@@

ggsave(filename = glue('{CACHE_DIR}/CRC_labs_raw_data_distribution_by_rel_date.png'), plot = p, width = 8, height = 6)
```

![CRC labs](../../cache/CRC_labs_raw_data_distribution_by_rel_date.png)

Use -180 as the time cutoff and separate the patient journey into two sections: 0 - 180 days before the first diagnosis of CRC, and >180 days before the first diagnosis of CRC. Within each section, we will pick the latest lab measurement for each LOINC lab test (features) and assign either + (within 180 days) or - outcome (outside of 180 days). 
```{r}
# prepare data entries for + and - outcomes
# each lab result is associated with an outcome
# briefly, the following snippet uses the following algorithm:
# calculate the relative date to the first diagnosis: if before diagnosis, - days otherwise + days
# then cut the relative days into a few windows (-Inf, -180), (-180, 0) and (0, Inf)
# for each window, we only use the closest one to diagnosis date as a representative, although we can use other methods such as the mean

@@

Skip the following sections.
<!-- ```{r max.output=-1, message=FALSE} -->
<!-- require(furrr) -->
<!-- plan(multisession, workers = 5) -->

<!-- crc_analysis <- df_labs_and_outcomes %>% -->
<!--   group_by(loinc) %>% -->
<!--   nest() %>% -->
<!--   ungroup() %>% -->
<!--   mutate(result = future_map(data, pipeline_compare_for_outcome_prediction)) %>% -->
<!--   select(-data) %>% -->
<!--   unnest(result) -->

<!-- crc_analysis -->
<!-- ``` -->

<!-- ```{r message=FALSE} -->
<!-- df_one_loinc <- df_labs_and_outcomes %>% filter(loinc == '4544-3') -->

<!-- df_hpo_features <- create_hpo_features(df_one_loinc = df_one_loinc) -->

<!-- df_train_test <- train_test_split(df_hpo_features, p = 0.7, list=FALSE) -->

<!-- df_train <- df_train_test$df_train -->
<!-- df_test <- df_train_test$df_test -->

<!-- lr_L <- lr_for_hpo_L(df_train = df_train, df_test = df_test) -->
<!-- roc_comparison <- compare_rocs(lr_L[[5]], lr_L[[6]]) -->
<!-- roc_comparison <- bind_cols(roc_comparison) -->

<!-- plot(lr_L[[5]]) -->
<!-- plot(lr_L[[6]]) -->

<!-- roc.test(lr_L[[5]], lr_L[[6]], method = "bootstrap") -->
<!-- ``` -->

@@

Skip the following code sections.

<!-- ```{r max.output=-1, message=FALSE} -->
<!-- require(furrr) -->
<!-- plan(multisession, workers=8) -->
<!-- start_time <- Sys.time() -->

<!-- crc_separate_controls_analysis <- df_labs_and_outcomes_case_control %>% -->
<!--   group_by(loinc) %>% -->
<!--   nest() %>% -->
<!--   ungroup() %>% -->
<!--   mutate(result = future_map(data, pipeline_compare_for_outcome_prediction, .options = furrr_options(seed = T))) %>% -->
<!--   select(-data) %>% -->
<!--   unnest(result) -->

<!-- end_time <- Sys.time() -->

<!-- print(end_time - start_time) -->

<!-- crc_separate_controls_analysis -->
<!-- ``` -->

<!-- ```{r message=FALSE} -->
<!-- df_one_loinc <- df_labs_and_outcomes_case_control %>% filter(loinc == '4544-3') -->

<!-- df_hpo_features <- create_hpo_features(df_one_loinc = df_one_loinc) -->

<!-- df_train_test <- train_test_split(df_hpo_features, p = 0.7, list=FALSE) -->

<!-- df_train <- df_train_test$df_train -->
<!-- df_test <- df_train_test$df_test -->

<!-- lr_L <- lr_for_hpo_L(df_train = df_train, df_test = df_test) -->
<!-- roc_comparison <- compare_rocs(lr_L[[5]], lr_L[[6]]) -->
<!-- roc_comparison <- bind_cols(roc_comparison) -->

<!-- plot(lr_L[[5]]) -->
<!-- plot(lr_L[[6]]) -->

<!-- roc.test(lr_L[[5]], lr_L[[6]], method = "bootstrap") -->
<!-- ``` -->

@@

p <- ggplot(df_demographics_with_outcomes, aes(x = outcome, y = propensity_score)) +
  geom_violin() +
  theme_az()

ggsave(glue("{CACHE_DIR}/plot_CRC_before_matching.png"), width = 2, height = 1.5)
p
```
```{r}
require(MatchIt)

@@

# try this stackoverflow method
# https://stackoverflow.com/questions/29672088/speedup-matchit
# still slow, but works
# df_demographics_with_outcomes %>% colnames()

run_matchit <- function(filepath, force_rerun) {
  if (force_rerun | !file.exists(filepath)){
    # place m.out in the global environment with "<<-" (instead of regular "<-")
    m.out <<- matchit(outcome ~ medical_record_number, data = df_demographics_with_outcomes, method = "nearest", distance = df_demographics_with_outcomes$propensity_score)
    # save the matchit object in cache
    save(m.out, file=filepath)
  }
  load(file=filepath)
}

run_matchit(filepath = glue('{CACHE_DIR}/m.out.obj'), force_rerun = FALSE)

matched_unique_ids_etc <- match.data(m.out, data = df_demographics_with_outcomes)

#matched_unique_ids_etc %>% group_by(outcome) %>% summarise(n =n())
p <- ggplot(matched_unique_ids_etc, aes(x = outcome, y = propensity_score)) +
  geom_violin() + 
  theme_az()

ggsave(glue("{CACHE_DIR}/plot_CRC_after_matching.png"), plot = p, width = 2, height = 1.5)

plot_match_age <- ggplot(matched_unique_ids_etc, aes(x = outcome, y = age, fill = outcome)) +
  geom_violin(draw_quantiles = c(0.5)) +
  theme(legend.position = 'none') + 
  theme_az()

plot_match_gender <- ggplot(matched_unique_ids_etc, aes(x = outcome, fill = gender)) +
  geom_bar(stat = 'count') + 
  theme_az()

plot_match_race <- ggplot(matched_unique_ids_etc, aes(x = outcome, fill = race)) +
  geom_bar(stat = 'count')+ 
  theme_az()

plot_match <- gridExtra::arrangeGrob(plot_match_age, plot_match_gender, plot_match_race, layout_matrix = rbind(c(1,2), c(3,3)), ncol = 1)

ggsave(filename = glue('{CACHE_DIR}/plot_CRC_age_gender_race_after_matching.png'), plot = plot_match, width = 6, height = 5)
plot(plot_match)
```
We need to put back the identifiers of controls to the database in order to selectively retrieve their labs. 
```{r}
dbExecute(dsca, "create temp table if not exists matched_control_cohort (medical_record_number VARCHAR)")

@@

# insert mrns one by one
matched_unique_ids_etc %>% filter(outcome == 'normal') %>% 
  rowwise() %>%
  mutate(batch_number = sample(seq(10), size = 1, replace = TRUE)) %>%
  ungroup() %>%
  group_by(batch_number) %>%
  summarise(sql_query = str_c(str_c("('", medical_record_number, "')"), collapse = ",")) %>% 
  ungroup() %>%
  mutate(sql_query = walk(sql_query, insert_mrns, dbconn = dsca))

# retrieve labs for the matched control cohort
sql <- read_file('../sql/get_labs_for_colorectal_cancer_matched_control.sql')
df_matched_control_labs <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = 'labs_colorectal_cancer_matched_controls.csv', db_access_callback = dbGetQuery, conn = dsca, statement = sql, force_rerun = FORCE_RERUN)
```

@@

```{r message=FALSE, max.output=-1}
require(furrr)
plan(multisession, workers=8)
start_time <- Sys.time()

@@

crc_case_matched_controls_analysis %>% write.csv(file = glue('{CACHE_DIR}/crc_case_matched_controls_analysis.csv'), row.names = FALSE)

crc_case_matched_controls_analysis
```

<!-- Combine demographics to see whether the AUC can be improved. It is not used in the manuscript.  -->
<!-- ```{r} -->
<!-- df_demographics_case_matched_control <- bind_rows(df_case_demographics %>% select(-min_diag_date),  -->
<!--                                                   df_control_demographics %>% filter(medical_record_number %in%  -->
<!--                                                                                        unique(df_matched_control_labs$medical_record_number)) %>% -->
<!--                                                     filter(!(medical_record_number %in% df_case_demographics$medical_record_number))) -->

<!-- plan(multisession, workers=4) -->
<!-- start_time <- Sys.time() -->

<!-- crc_case_matched_controls_analysis_with_demographics <- df_labs_and_outcomes_case_matched_control %>% -->
<!--   left_join(df_demographics_case_matched_control, by = 'medical_record_number') %>% -->
<!--   mutate(age = as.duration(interval(date_of_birth, lab_date)) / dyears()) %>% -->
<!--   filter(!(race %in% c('African American', 'American Indian or Alaska Native', 'Native Hawaiian or Other Pacific Islander'))) %>% -->
<!--   # mutate(gender = as.factor(gender), -->
<!--   #        race = as.factor(race)) %>%  -->
<!--   # mutate(gender = relevel(gender, 'Male'), -->
<!--   #        race = relevel(race, 'White')) %>% -->
<!--   group_by(loinc) %>% -->
<!--   nest() %>% -->
<!--   ungroup() %>%  -->
<!--   mutate(result = future_map(data, pipeline_compare_for_outcome_prediction, include_demographics=TRUE, .options = furrr_options(seed = T))) %>% -->
<!--   select(-data) %>% -->
<!--   unnest(result) -->

<!-- end_time <- Sys.time() -->

<!-- print(end_time - start_time) -->

<!-- crc_case_matched_controls_analysis_with_demographics -->
<!-- ``` -->

### 2022-04-26 code/r/02_E_numeric_vs_hpo_combined.Rmd
update
continue refactoring

---
title: "numeric vs HPO"
output: html_notebook
---

@@

plot_AUC_differences_numeric_vs_hpo <- ggplot(df, aes(x = method, y = auc, group = interaction(disease, loinc))) + 
  geom_point(size = 1) +
  geom_line(aes(color = disease)) +
  scale_x_discrete(breaks = c('continuous', 'hpo'), labels = c('original numeric value', 'transformed to HPO')) +
  xlab("Use Lab Test Results As") +
  ylab("AUC") +
  theme_bw() + 
    theme(panel.grid = element_blank(), legend.position = 'none',
          axis.text = element_text(size = 8),
          axis.title = element_text(size = 8), 
          strip.text.x = element_text(margin = margin(0.3, 0, 0.3, 0), size = 5),
          plot.title = element_text(size = 8)) +
  facet_wrap(~ disease)

ggsave(filename = glue('{CACHE_DIR}/plot_AUC_differences_numeric_vs_hpo.pdf'), width = 6, height = 6)

>>>>

---
title: "numeric vs HPO"
output:
  html_document:
    df_print: paged
---

@@

plot_AUC_differences_numeric_vs_hpo <- ggplot(df, aes(x = method, y = auc, group = interaction(disease, loinc))) + 
  geom_point(size = 1) +
  geom_line(aes(color = disease)) +
  scale_x_discrete(breaks = c('continuous', 'hpo'), labels = c('original numeric value', 'transformed to HPO')) +
  xlab("Use Lab Test Results As") +
  ylab("AUC") +
  theme_bw() + 
    theme(panel.grid = element_blank(), legend.position = 'none',
          axis.text = element_text(size = 8),
          axis.title = element_text(size = 8), 
          strip.text.x = element_text(margin = margin(0.3, 0, 0.3, 0), size = 8),
          plot.title = element_text(size = 8)) +
  facet_wrap(~ disease)

ggsave(filename = glue('{CACHE_DIR}/plot_AUC_differences_numeric_vs_hpo.pdf'), width = 6, height = 5)
